---
title: "The Art of Being Systematically Less Wrong"
---

# The Art of Being Systematically Less Wrong {#sec-motivation}

There are two kinds of forecasters: those who can't predict the future,
and those who don't know they can't predict the future.

The first group has an advantage. They stop trying to be right and start
trying to be *less wrong* — and that turns out to be a problem you can
actually solve.

---

## A decision you make every day {#sec-motivation-decision}

You are running out of coffee at home. You need to decide how many bags
to buy. Buy too few and you run out before the week ends. Buy too many
and they go stale before you finish them. Neither is catastrophic, but
both have a cost.

Without thinking about it, you are solving an optimization problem. You
have a loss — the combined cost of under-buying and over-buying — and you
are trying to minimize it given what you know: your consumption rate, how
long bags last, how often you go to the store.

Now scale that up. You are not buying coffee for one person. You are
managing the electricity grid for an entire region. Every thirty minutes,
you need to decide how much power to procure for the next interval. Buy
too little and you impose rolling blackouts. Buy too much and you pay for
energy that goes nowhere. The cost of being wrong is not stale coffee —
it is millions of dollars and compromised infrastructure.

The structure of the problem is the same. The stakes are not.

---

## Two datasets, one question {#sec-motivation-datasets}

This book uses two datasets as its main thread. They come from different
domains and have different structures, but they share the same underlying
question: *given what you can observe, what is your best prediction of
what you cannot?*

### Victorian electricity demand {#sec-motivation-vicelec}

`vic_elec` records half-hourly electricity demand and temperature across
the state of Victoria, Australia, from 2012 to 2014. Every row is a
moment. Every moment is a decision someone had to make about how much
power to have available.

```{r}
#| label: fig-vicelec-overview
#| fig-cap: "Half-hourly electricity demand (top) and temperature (bottom) in Victoria, Australia, January 2014. The relationship is clear but imprecise — temperature alone does not determine demand."
#| message: false
#| warning: false

library(tsibbledata)
library(dplyr)
library(ggplot2)
library(patchwork)
library(here)

vic_elec <- readr::read_csv(here("data", "vic_elec.csv"),
                             show_col_types = FALSE) |>
  mutate(Time = as.POSIXct(Time))

vic_sample <- vic_elec |>
  filter(Time >= as.POSIXct("2014-01-01"),
         Time <  as.POSIXct("2014-01-15"))

p_demand <- ggplot(vic_sample, aes(x = Time, y = Demand)) +
  geom_line(colour = "#2166ac", linewidth = 0.5) +
  labs(x = NULL, y = "Demand (MW)") +
  theme_minimal(base_size = 11)

p_temp <- ggplot(vic_sample, aes(x = Time, y = Temperature)) +
  geom_line(colour = "#d6604d", linewidth = 0.5) +
  labs(x = NULL, y = "Temperature (°C)") +
  theme_minimal(base_size = 11)

p_demand / p_temp
```

Temperature drives demand — air conditioning is the dominant load in an
Australian summer. But the relationship shifts. A hot Tuesday looks
different from a hot Saturday. A heat wave on the third consecutive day
draws more than a heat wave on the first. The signal is real; it is just
not simple enough to write down by hand.

::: {.panel-tabset}

## R

```{r}
#| label: tbl-vicelec-glimpse
#| message: false

vic_elec |>
  select(Time, Demand, Temperature, Holiday) |>
  slice_head(n = 8) |>
  knitr::kable(digits = 2,
               caption = "First eight rows of `vic_elec`.")
```

## Python

```{python}
#| label: tbl-vicelec-glimpse-py

from pathlib import Path
import pandas as pd

def _project_root(markers=("_quarto.yml", "renv.lock")):
    for d in (Path.cwd(), *Path.cwd().parents):
        if any((d / m).exists() for m in markers):
            return d
    return Path.cwd()

vic_elec = pd.read_csv(_project_root() / "data" / "vic_elec.csv")  # <1>
vic_elec[["Time", "Demand", "Temperature", "Holiday"]].head(8)
```
1. Loaded from the book's `data/` folder; original source is the
   `tsibbledata` R package (half-hourly Victorian electricity demand,
   2012–2014).

:::

### California housing prices {#sec-motivation-calhousing}

The second dataset is 20,640 California census blocks from the early
1990s. Each block is described by eight numbers: median income, house
age, average rooms, average bedrooms, population, average occupancy,
latitude, longitude. The target is median house value.

```{r}
#| label: fig-calhousing-overview
#| fig-cap: "Median house value across California census blocks. Coastal proximity matters, but income, density, and other variables all play a role. No single feature explains the target."
#| message: false
#| warning: false

library(ggplot2)
library(here)

California <- readr::read_csv(here("data", "california_housing.csv"),
                               show_col_types = FALSE)

ggplot(California, aes(x = longitude, y = latitude,
                       colour = log(median_house_value))) +
  geom_point(size = 0.3, alpha = 0.5) +
  scale_colour_viridis_c(option = "plasma", name = "log(value)") +
  coord_fixed(ratio = 1.3) +
  labs(x = "Longitude", y = "Latitude") +
  theme_minimal(base_size = 11)
```

There is no time axis here. Census blocks do not arrive in sequence —
each one is independent. What matters is not what happened before, but
which other blocks are *similar* and whether similar blocks tend to have
similar prices. The coastal premium is visible in the plot. So is the
concentration of high-value blocks in the Bay Area and Los Angeles. But
the relationship between the eight features and the target is not
something you can read off a map.

::: {.panel-tabset}

## R

```{r}
#| label: tbl-calhousing-glimpse
#| message: false

California |>
  slice_head(n = 8) |>
  knitr::kable(digits = 2,
               caption = "First eight rows of the California Housing dataset.")
```

## Python

```{python}
#| label: tbl-calhousing-glimpse-py

from sklearn.datasets import fetch_california_housing  # <1>
import pandas as pd

cal = fetch_california_housing(as_frame=True)
cal.frame.head(8)
```
1. `fetch_california_housing()` returns the same dataset from
   scikit-learn. Note: `MedHouseVal` is in units of \$100,000.

:::

---

## Why a formula is not enough {#sec-motivation-formula}

When you write `=0.3 * B2 + 85000` in a spreadsheet, you have already
made the decision. You chose the slope. You chose the intercept. The
formula does what you told it to do, nothing more.

For the electricity problem, you do not know the slope in advance. You do
not even know that the relationship is linear. The only honest starting
point is the data itself — and the question of what it means for a
prediction to be good.

That question has a precise answer. You choose a way to measure how wrong
a prediction is. Then you find the function that is least wrong, on
average, over all the data you have. That search is optimization. Every
model in this book is the result of that search, with different choices
about what "wrong" means and what functions are allowed.

---

## What comes next {#sec-motivation-next}

Part I builds the optimization machinery from scratch. You will see how
to find the minimum of a function when there are no constraints, then
when there are. You will learn to read the conditions that tell you
whether a solution is optimal. By the end, you will have the tools to
derive models from first principles — not just use them.

Parts II and III introduce the models themselves: how to measure
similarity between observations, and how to use that similarity to build
predictions that generalize beyond the training data.

The final chapters return to these two datasets with models that optimize
percentage error directly — a choice that matters when a \$50,000
prediction error means something very different for a \$100,000 house
than for a \$600,000 house.

---

## Exercises {#sec-motivation-exercises}

::: {.callout-note collapse="true"}
### Exercise 1.1 — Reading the time series

Load `vic_elec` and compute the average half-hourly demand separately
for weekdays and weekends. Plot both on the same axis. At what hour of
the day is the gap between weekday and weekend demand largest?
:::

::: {.callout-note collapse="true"}
### Exercise 1.2 — The cost of being wrong

Suppose electricity costs \$150 per MWh during peak hours (5–9 PM on
weekdays) and \$40 per MWh at all other times. Using `vic_elec`:

1. Compute the total procurement cost if you always buy exactly what
   was demanded.
2. Compute the cost if your forecast is 10% high during peak hours
   (you over-procure; the excess is wasted).
3. Compute the cost if your forecast is 10% low during peak hours
   (you under-procure; you must buy the deficit at a spot premium of
   \$500 per MWh).

Which direction of error is more expensive? Does your answer change
how you would define a good prediction?
:::

::: {.callout-note collapse="true"}
### Exercise 1.3 — What does "wrong" mean?

For the California Housing data, compute prediction errors for two
naive models: one that always predicts the training mean, and one that
always predicts \$200,000.

Compute MAE, RMSE, and MAPE for each. Which metric best captures
whether the model is useful? Does a \$50,000 error mean the same thing
for a \$100,000 house and a \$600,000 house?
:::

::: {.callout-note collapse="true"}
### Exercise 1.4

A loss function $\ell(\hat{y}, y)$ is *scale-free* if multiplying both
the prediction and the target by the same positive constant leaves the
loss unchanged: $\ell(c\hat{y}, cy) = \ell(\hat{y}, y)$ for all
$c > 0$.

Show that squared error $(\hat{y} - y)^2$ is not scale-free. Then
construct a loss function that is. Is it unique? What structural
property do all scale-free loss functions share?
:::
