---
title: "From Spreadsheet to Function"
---

# From Spreadsheet to Function {#sec-motivation}

What do you do when a spreadsheet is no longer enough?

Not because the data is too large. Not because the formulas are too slow.
Because the question you are asking cannot be answered with a formula you
can write by hand — and you know it.

This is where most data work actually starts. Not with a model, not with an
algorithm. With a question that a pivot table cannot answer, and the
uncomfortable realization that something more systematic is needed.

This book is about building that something from scratch.

---

## A company's electricity bill, by the half-hour {#sec-motivation-vicelec}

Consider a company that operates in Victoria, Australia. Its infrastructure
runs on electricity — servers, cooling systems, office equipment. Every
half-hour, the grid records how much power the company draws. Every
half-hour, the wholesale price of that electricity changes based on how
much everyone else is drawing at the same time.

If you can predict demand before it happens, you can act: pre-scale
servers, shift workloads, negotiate contracts. If you cannot, you react —
and reacting is expensive.

```{r}
#| label: fig-vicelec-overview
#| fig-cap: "Victorian electricity demand and temperature over a two-week window. Demand tracks temperature, but not perfectly — and the gap between them is where the forecasting problem lives."
#| message: false
#| warning: false

library(here)
library(readr)
library(dplyr)
library(ggplot2)

vic_elec <- read_csv(here::here("data", "vic_elec.csv"))  # <1>

vic_sample <- vic_elec |>  # <2>
  filter(Time >= as.POSIXct("2014-01-01"),
         Time <  as.POSIXct("2014-01-15"))

ggplot(vic_sample, aes(x = Time)) +
  geom_line(aes(y = Demand, colour = "Demand (MW)"),
            linewidth = 0.5) +
  geom_line(aes(y = Temperature * 30, colour = "Temperature (°C × 30)"),
            linewidth = 0.5, linetype = "dashed") +
  scale_colour_manual(
    values = c("Demand (MW)" = "#2166ac",
               "Temperature (°C × 30)" = "#d6604d")
  ) +
  labs(x = NULL, y = "Demand (MW)", colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```
1. Loaded from the book's `data/` folder; original source is the `tsibbledata` R package (half-hourly Victorian electricity demand, 2012–2014).
2. We filter to January 2014 — peak summer in Australia — where demand is highest and most variable.

The pattern is visible but imprecise. Temperature drives demand — air
conditioning is the dominant load — but the relationship shifts with the
day of the week, with holidays, with whether it was already hot the day
before. A line through a scatter plot will not capture that. A lookup
table will not generalize to next week.

What we need is a *function* — one that takes what we know now (temperature,
hour, day type) and returns a prediction we can act on.

::: {.panel-tabset}

## R

```{r}
#| label: tbl-vicelec-glimpse
#| message: false

library(readr)

vic_elec <- read_csv(here::here("data", "vic_elec.csv"))

vic_elec |>
  select(Time, Demand, Temperature, Holiday) |>
  slice_head(n = 8) |>
  knitr::kable(digits = 2,
               caption = "First eight rows of `vic_elec`.")
```

## Python

```{python}
#| label: tbl-vicelec-glimpse-py
#| message: false

import pandas as pd
from pathlib import Path

def _project_root(markers=("_quarto.yml", "renv.lock")):
    for d in (Path.cwd(), *Path.cwd().parents):
        if any((d / m).exists() for m in markers):
            return d
    return Path.cwd()

vic_elec = pd.read_csv(_project_root() / "data" / "vic_elec.csv")  # <1>

(
  vic_elec[["Time", "Demand", "Temperature", "Holiday"]]
  .head(8)
)
```
1. Loaded from the book's data folder; original source is the tsibbledata R package (half-hourly Victorian electricity demand, 2012–2014).

:::

The dataset has `r nrow(vic_elec)` rows — one per half-hour from 2012 to
2014. Every row is a moment when someone, somewhere, made a decision about
how much electricity to use. Together, they describe a system with memory,
seasonality, and noise.

---

## House prices across California {#sec-motivation-calhousing}

The second running example is quieter. No time index, no recurring pattern.
Just 20,640 census blocks from California in the early 1990s, each described
by eight numbers: median income, house age, average rooms, average bedrooms,
population, average occupancy, latitude, longitude. The target is median
house value.

```{r}
#| label: fig-calhousing-overview
#| fig-cap: "Median house value across California census blocks, plotted by geographic location. Values are compressed at the upper end — a pattern that makes percentage error a more natural metric than raw dollar error."
#| message: false
#| warning: false

library(readr)  # <1>
library(ggplot2)

California <- read_csv("data/california_housing.csv")  # <2>

ggplot(California, aes(x = longitude, y = latitude,
                       colour = log(median_house_value))) +
  geom_point(size = 0.4, alpha = 0.6) +
  scale_colour_viridis_c(option = "plasma",
                         name = "log(value)") +
  coord_fixed(ratio = 1.3) +
  labs(x = "Longitude", y = "Latitude") +
  theme_minimal(base_size = 12)
```
1. The California Housing data ships as a CSV in the book's `data/` folder; it is the same dataset scikit-learn exposes via `fetch_california_housing()`.
2. The full dataset has 20,640 rows. We use it as-is here; experiments later use a stratified subsample for speed.

::: {.panel-tabset}

## R

```{r}
#| label: tbl-calhousing-glimpse
#| message: false

library(readr)

California <- read_csv("data/california_housing.csv")

California |>
  slice_head(n = 8) |>
  knitr::kable(digits = 2,
               caption = "First eight rows of the California Housing dataset.")
```

## Python

```{python}
#| label: tbl-calhousing-glimpse-py

import pandas as pd
from sklearn.datasets import fetch_california_housing  # <1>

cal = fetch_california_housing(as_frame=True).frame
cal = cal.rename(columns={  # <2>
    "MedInc": "median_income", "HouseAge": "house_age",
    "AveRooms": "average_rooms", "AveBedrms": "average_bedrooms",
    "Population": "population", "AveOccup": "average_occupancy",
    "Latitude": "latitude", "Longitude": "longitude",
    "MedHouseVal": "median_house_value",
})
cal["median_house_value"] = cal["median_house_value"] * 100_000  # <3>
cal.head(8)
```
1. `fetch_california_housing()` returns the same data scikit-learn ships.
2. We rename to the book's snake_case columns so the table matches the R tab.
3. scikit-learn reports `MedHouseVal` in units of \$100,000; we rescale to raw dollars to match `data/california_housing.csv`.

:::

The prediction problem here is different in character from the electricity
problem. There is no time axis. Each observation is independent — knowing
the house value in Sacramento tells you nothing about the next row. What
matters is the *relationship* between the eight features and the target.
And that relationship is not linear. Census blocks near the coast behave
differently from inland blocks even when income is held fixed. The geography
matters in ways that do not reduce to a single variable.

Again: a formula you can write by hand will not capture this. You need a
function that learns the pattern from the data.

---

## What functions do that equations do not {#sec-motivation-functions}

When you write a formula in a spreadsheet — say, `=0.5 * B2 + 1200` — you
are making a decision. You chose the slope. You chose the intercept. The
formula does exactly what you told it to do, no more.

That is fine when the relationship is simple and known in advance. It fails
when the relationship is complex, nonlinear, and discovered only by looking
at the data.

This book is about the transition from formulas you write to functions that
are *fitted* — determined by an optimization problem rather than by hand.
The core idea is simple, even if the machinery is not: you describe what
a good prediction looks like (a loss function), and you search for the
function that makes predictions as good as possible (optimization).

Everything else — kernels, dual problems, support vectors — is in service
of this idea.

---

## Two problems, one framework {#sec-motivation-oneframework}

These two datasets are not chosen at random. They represent the two main
structural types of prediction problem that appear in practice.

The electricity data is a *time series*: observations are ordered in time,
and the recent past matters. Predicting tomorrow's demand depends on what
happened yesterday. Patterns repeat weekly and annually. The right model
must respect time.

The California data is *cross-sectional*: observations are exchangeable,
and there is no natural ordering. What matters is the geometry of the
feature space — which inputs are similar, and whether similar inputs produce
similar outputs.

The models developed in this book handle both. The optimization framework
is the same. The loss function adapts to the measurement scale. The kernel
decides what "similar" means.

By Part IV, you will have the tools to fit these datasets with SVR models
that optimize percentage error directly — a choice that matters when the
target varies over several orders of magnitude, as house prices and
electricity demand both do.

For now, look at the data. Ask what it would mean to predict it well. The
answer to that question will motivate every chapter that follows.

---

## Exercises {#sec-motivation-exercises}

::: {.callout-note collapse="true"}
### Exercise 1.1 — Reading the time series

Load `vic_elec` and compute the average half-hourly demand separately for
weekdays and weekends. Plot both on the same axis. At what time of day is
the gap between weekday and weekend demand largest?

*What you need:* `dplyr`, `lubridate`, `ggplot2`.
:::

::: {.callout-note collapse="true"}
### Exercise 1.2 — The price of a bad prediction

Suppose electricity costs \$150 per MWh during peak hours (5–9 PM weekdays)
and \$40 per MWh otherwise. Using the `vic_elec` data:

1. Compute the total cost if you always procure exactly what was demanded.
2. Compute the cost if your forecast is systematically 10% high during peak
   hours (you over-procure and the excess is wasted at zero value).
3. Compute the cost if your forecast is 10% low during peak hours (you
   under-procure and must buy the deficit at a spot premium of \$500/MWh).

Which direction of error is more expensive? Does this change how you would
define "a good prediction"?
:::

::: {.callout-note collapse="true"}
### Exercise 1.3 — Scale and error metrics

For the California Housing data, compute three prediction errors for a
naive model that predicts the training mean for every observation:

- **MAE** — mean absolute error (in dollars)
- **RMSE** — root mean square error (in dollars)
- **MAPE** — mean absolute percentage error (unitless)

Now repeat for a model that always predicts $200,000.

Which error metric is most informative? Does a \$50,000 error mean the
same thing for a \$100,000 house as for a \$600,000 house?

*This question motivates Chapter 16.*
:::

::: {.callout-note collapse="true"}
### Exercise 1.4 — Graduate level: loss functions as design choices

The choice of loss function is a statement about what errors matter.

Consider a delivery company predicting package transit times. A 2-hour
error on a 4-hour delivery is qualitatively different from a 2-hour error
on a 48-hour delivery — yet mean squared error treats them identically.

Formalize this argument. Define a loss function $\ell(\hat{y}, y)$ that
satisfies: (a) it is zero when prediction is perfect, (b) it is scale-free
(invariant to multiplying both $\hat{y}$ and $y$ by a positive constant),
and (c) it penalizes over-prediction and under-prediction symmetrically.

Is there more than one such function? What structural property do they
share?
:::
